A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-25 paper.tex
Minor rephrasing to reduce word count
And also to introduce the idea of local trees early on.

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately, however, there is substantial confusion over
terminology and little agreement on shared standards for data interchange,
significantly hampering progress in this important area.
This confusion and lack of agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a stochastic process (the coalescent with recombination) together with its
representation, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individual's monoploid genomes, an edge
% jk-note: this is clumsy
defines a genetic ancestry relationship between two nodes,
and annotations on those edges define the patterns of
inheritance in intervals along the genome.
We show how this definition can be used to generalise the encoding to
cover both the standard and several simplified versions of ARGs, as well as
enabling efficient access to local trees along the genome. Additionally, we
define some classifications and metrics of ARG nodes. Finally, we relate
these generalisations to the ARGs produced by a number of recently
developed ancestral inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

@@

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of
species~\citep{rannala2003genetics}, non-recombining sections of
DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
and viruses~\citep{grenfell2004science}.
A genealogical tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the
trees~\citep{tajima1983evolutionary,tavare1984line}
% Do we mean "despite the fact that analysis of genetic data relies on methods
% which ultimately derive from the inheritance paths defined by genetic trees
% and the mutations upon them, analysis so far has focussed on summaries ...
such as allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical
genetics.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies, leading to basic errors in statements
about what different inference methods produce as output, as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

\section*{Ancestral graphs: a brief history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
The coalescent is a stochastic process, where each random realisation
is a genealogical tree describing the genetic
ancestors of the samples and their inheritance relationships.
A great strength of the coalescent as a modelling framework
is its flexibility, and numerous elaborations have been
incorporated~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
In the first extension to the basic process, \cite{hudson1983properties}
analysed the effects of recombination on the random genealogical
trees generated by the coalescent,
producing several fundamental analytical results
as well as describing the elegant algorithm
at the heart of simulation
programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}
used in thousands of studies.
In its original formulation, Hudson's algorithm produced a
sequence of genealogical trees along the genome.
Critically, although the correlation structure among these
local trees was analysed analytically, those correlations
were not captured explicitly: the ancestry of the sample
was viewed as a series of disconnected local trees rather than
as a single cohesive object.

@@

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
It is important to note that such recombination-parsimony ARGs
have quite different properties from realisations of the
(previously synonymous) ARG \emph{process}. Firstly, branch lengths are
not typically estimated, and node times are required
order to compute a likelihood under the coalescent with recombination
(see section XXX). Secondly, even if branch lengths were estimated,
the realisations would have a very small likelihood since
the coalescent with recombination is often highly \emph{un}parsimonious
in terms of recombination events for non-trivial scaled recombination rate.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was now therefore broken.
\cite{minichiello2006mapping} made this separation of the ARG as a
stochastic process vs the ARG as a data structure explicit, and
argued for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep{mathieson2020ancestry,brandt2021evaluation}
[TODO expand sentence].
Gusfield~\citep{gusfield2014recombinatorics}
defines also defines an ARG as a concrete realisation
and studies instances that satisfy particular
properties such as ``galled trees''
\citep{wang2001perfect, gusfield2004optimal}
as well as describing algorithms for more general phylogenetic
networks~\citep{huson2010phylogenetic}.

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
% Jere: ASG section redrafted, gene transfer left as TODO (because I
% know essentially nothing about it).
In parallel with this tendancy to decouple the term
Ancestral Recombination Graph from its original defining stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode desired
patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
Similarly, the Ancestral Gene Transfer Graph

@@

The point of this historical overview is to illustrate that the term
Ancestral Recombination Graph has different meanings in different
contexts, and in its original form it was closely coupled to
the coalescent with recombination. In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely derived from the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events.
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occured in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we refer to it as an ``event ARG``, or eARG for brevity.
An example of this form of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an ARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

% What do we mean by an eARG, formally?
To be concrete, we can formally define the
classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(This formulation is equivalent to having different ``types''
of node, and that breakpoints are associated with recombination
nodes only).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
This encoding of an eARG is equivalent to the output
produced by programs such as
\texttt{ARGweaver}~\citep{rasmussen2014genome}
and \texttt{KwARG}~\citep{ignatieva2021kwarg},
and example is given in Fig.~\ref{fig-arg-data-structure}A.
Note that the description above captures only the
graph topology: if we also wish to know the branch lengths we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG (the most basic downstream
operation for an ARG data structure).
To recover the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event above a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% TODO update this when then later sections have been drafted.
As we see in sections XXX and YYY there are substantial limitations
of expressivity and computational performance when using this eARG
encoding. More fundamentally, however, the general approach
of classifying nodes according to a discrete set of events
and associating genome information with those nodes is intrinsically
limiting. As we saw in the previous section, the original ARG
definitions were derived directly from the coalescent with recombination
stochastic process, and the types of ancestry that can be
described are therefore limited by the assumptions of that
model. While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.

@@

Another assumption of the coalescent with recombination which prevents complex
combinations of events occuring simultaneously, is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
Clearly multiple recombinations (and other complex events) will
occur in the ancestry of such datasets, and we should not be
limited in what we can infer by the data structure that we
use to represent the results.

>>>>

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately confusion exists over terminology and there is
little agreement on shared standards for data interchange,
significantly hampering progress in the field.
This lack of clarity and agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a stochastic process (the coalescent with recombination) together with its
representation, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, 
% is it ambiguous? How? Can we simply delete this?
fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individual's monoploid genomes and an edge
defines genetic inheritance between two nodes, with annotations on
each edge specifying which regions of the genome have
been inherited via that route.
We show how this definition can be used to generalise the encoding to
cover both the standard and several simplified versions of ARGs, as well as
enabling efficient access to local trees along the genome. Additionally, we
define some classifications and metrics of ARG nodes. Finally, we relate
these generalisations to the ARGs produced by a number of recently
developed ancestral inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

@@

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary or phylogenetic tree is a necessary prerequisite for many analyses of
species~\citep{rannala2003genetics}, non-recombining sections of
DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
and viruses~\citep{grenfell2004science}.
A phylogenetic tree captures information about the genealogy
of the sample in an elegant and concise
way, by postulating a set of ancestors common to the samples
together with paths of inheritance between them. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}.
However, the genetic genealogy of a recombining organism
cannot be described as a single evolutionary tree: instead,
the inheritance of each local section of the genome may be described
by a (slightly) different tree~\citet{hudson1983properties}. Inferring complex
genealogies of this nature
presents profound technical difficulties, meaning tree-based approaches are
rarely used in population and statistical genetics. Therefore until recently, analysis of
population genetic data has primarily focussed on summaries derived
from the local trees~\citep{tajima1983evolutionary,tavare1984line}
rather than using the underlying genealogy itself. Such
derived measures include allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
principal components~\citep{mcvean2009genealogical}.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which risks hampering
progress: we currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies. This leads to basic errors in statements
about what different inference methods produce as output, as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

\section*{Ancestral graphs: a brief history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
It is a stochastic process, where each random realisation
is a genealogical tree describing the genetic ancestry of the samples.
Numerous elaborations have been incorporated into the
basic process~\citep{hudson1990gene,hein2004gene,wakely2008coalescent};
in particular, early work by \citet{hudson1983properties}
analysed the effects of recombination on the random genealogical
trees generated by the coalescent. This
generated several fundamental analytical results
as well as describing the elegant algorithm
at the heart of simulation
programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}
used in thousands of studies.
In its original formulation, Hudson's algorithm produced a
sequence of genealogical trees along the genome.
Critically, although the correlation structure among these
local trees was analysed analytically, those correlations
were not captured explicitly: the ancestry
was viewed as a series of disconnected local trees rather than
as a single cohesive object.

@@

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
It is important to note that such recombination-parsimony ARGs
have quite different properties from realisations of the
ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
Firstly, branch lengths are not typically estimated,
% why do we need the following two lines
and node times are required in order to compute a likelihood under the
coalescent with recombination (see section XXX).
Secondly, even if branch lengths were estimated,
the realisations would have a very small likelihood since
the coalescent with recombination is often highly \emph{un}parsimonious
in terms of recombination events for non-trivial scaled recombination rate.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was now therefore broken.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep{mathieson2020ancestry,brandt2021evaluation}
[TODO expand sentence].
Gusfield~\citep{gusfield2014recombinatorics}
defines also defines an ARG as a concrete realisation
and studies instances that satisfy particular
properties such as ``galled trees''
\citep{wang2001perfect, gusfield2004optimal}
as well as describing algorithms for more general phylogenetic
networks~\citep{huson2010phylogenetic}.

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
% Jere: ASG section redrafted, gene transfer left as TODO (because I
% know essentially nothing about it).
In parallel with this tendency to decouple the term
Ancestral Recombination Graph from its original defining stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode desired
patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
Similarly, the Ancestral Gene Transfer Graph

@@

The point of this historical overview is to illustrate that the term
Ancestral Recombination Graph has different meanings in different
contexts, and in its original form it was closely coupled to
the coalescent with recombination. In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely derived from the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we refer to it as an ``event ARG``, or eARG for brevity.
An example of this form of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an ARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

% What do we mean by an eARG, formally?
To be concrete, we can formally define the
classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(This formulation is equivalent to having different ``types''
of node, with breakpoints are associated with recombination
nodes only).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
This encoding of an eARG is equivalent to the output
produced by programs such as
\texttt{ARGweaver}~\citep{rasmussen2014genome}
and \texttt{KwARG}~\citep{ignatieva2021kwarg}:
an example is given in Fig.~\ref{fig-arg-data-structure}A.
Note that the description above captures only the
graph topology: if we also wish to know the branch lengths we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes
rather than edges: we need to know from which parent the 
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG (the most basic downstream
operation for an ARG data structure).
To recover the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event above a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% TODO update this when later sections have been drafted.
As we see in sections XXX and YYY there are substantial limitations
of expressivity and computational performance when using this eARG
encoding. More fundamentally, however, the general approach
of classifying nodes according to a discrete set of events
and associating genome information with those nodes is intrinsically
limiting. As we saw in the previous section, the original ARG
definitions were derived directly from the coalescent with recombination
stochastic process, and the types of ancestry that can be
described are therefore limited by the assumptions of that
model. While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.

@@

Another assumption of the coalescent with recombination which prevents complex
combinations of events occurring simultaneously, is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
Clearly multiple recombinations (and other complex events) will
occur in the ancestry of such datasets, and we should not be
limited in what we can infer by the data structure that we
use to represent the results.
